How to \underline{not} write Virtual Machines for \mbox{Dynamic Languages}

### 2011-03-03 talk/dyla2007/talk.tex
arigo? This must be one of Armin's because I did not edit this file. But if I have instead gone and clobbered his changes with an earlier version a) I apologise and b) I have no clue how this could happen given that he checked in things later than I.

\date{3rd Workshop on Dynamic Languages and Applications, July 31 2007}

\pgfdeclareimage[height=0.5cm]{pypy-logo}{image/py-web.png}
\logo{\pgfuseimage{pypy-logo}}

@@

  \begin{itemize}
  \item
    implementing dynamic languages \par(with a focus on complicated ones)
  \item
    in a context of limited resources \par(academic, open source, or
    domain-specific)
  \end{itemize}
  \pause
  \begin{block}{
    Our point}
    \begin{itemize}
    \item
      Do not write virtual machines ``by hand''
    \item
      Instead, write interpreters in high-level languages
    \item
      Meta-programming is your friend
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Common Approaches to VM construction}
  \begin{block}{
    Using C directly (or C disguised as another language)}
    \begin{itemize}
    \item
      CPython
    \item
      Ruby
    \item
      Spidermonkey (Mozilla's JavaScript VM)
    \item
      but also: Squeak, Scheme48
    \end{itemize}
  \end{block}
  \begin{block}{
    Building on top of a general-purpose OO VM}
    \begin{itemize}
    \item
      Jython, IronPython
    \item
      JRuby, IronRuby
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Implementing VMs in C}
  When writing a VM in C it is hard to reconcile:
  \begin{itemize}
  \item
    flexibility, maintainability
  \item
    simplicity of the VM
  \item
    performance (needs dynamic compilation techniques)
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      \alert{CPython} is a very simple bytecode VM, performance not great
    \item
      \alert{Psyco} is a just-in-time-specializer, very complex, hard to
      maintain, but good performance
    \item
      \alert{Stackless} is a fork of CPython adding microthreads. It was never
      incorporated into CPython for complexity reasons
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Compilers are a bad encoding of Semantics}
  \begin{itemize}
  \item to reach good performance levels, dynamic compilation is often needed
  \item a dynamic compiler needs to encode language semantics
  \item this encoding is often obscure and hard to change
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      Psyco is a dynamic compiler for Python
    \item
      synchronizing with CPython's rapid development is a lot of effort
    \item
      many of CPython's new features not supported well
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Fixing of Early Design Decisions}
  \begin{itemize}
  \item when starting a VM in C, many design decisions need to be made upfront
  \item examples: memory management technique, threading model
  \item the decision is manifested throughout the VM source
  \item very hard to change later
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      CPython uses reference counting, increfs and decrefs everywhere
    \item
      CPython uses OS threads with one global lock, hard to change to
      lightweight threads or finer locking
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Implementation Proliferation}
  \begin{itemize}
  \item
    restrictions of the original implementation lead to re-implementations,
    forks
  \item
    all implementations need to be synchronized with language evolution
  \item
    lots of duplicate effort
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      several serious implementations: CPython, Stackless, Psyco, Jython,
      IronPython, PyPy
    \item
      the implementations have various grades of compliance
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Implementing Languages on Top of General-Purpose OO VMs}
  \begin{itemize}
  \item
    users wish to have easy interoperation with the general-purpose OO VMs used
    by the industry (JVM, CLR)
  \item
    therefore re-implementations of the language on the OO VMs are started
  \item
    even more implementation proliferation
  \item
    implementing on top of an OO VM has its own set of problems
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      \alert{Jython} is a Python-to-Java-bytecode compiler
    \item
      \alert{IronPython} is a Python-to-CLR-bytecode compiler
    \item
      both are slightly incompatible with the newest CPython version (especially
      Jython)
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Benefits of implementing on top of OO VMs}
  \begin{itemize}
  \item
    higher level of implementation
  \item
    the VM supplies a GC and mostly a JIT
  \item
    better interoperability than what the C level provides
  \item
    some proponents believe that eventually one single VM should be enough
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      both Jython and IronPython integrate well with their host OO VM
    \item
      XXX what else?
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{The problems of OO VMs}
  \begin{itemize}
  \item
    some of the benefits of OO VMs don't work out in practice
  \item
    most immediate problem: it can be hard to map
    concepts of the dynamic lang to
    the host OO VM
  \item
    performance is often not improved, and can be very bad, because of the
    semantic mismatch between the dynamic language and the host VM
  \item
    poor interoperability with everything outside the OO VM
  \item
    in practice, one OO VM is not enough
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      Jython about 5 times slower than CPython
    \item
      IronPython is about as fast as CPython (but some introspection features
      missing)
    \item
      XXX something about semantic mismatch?
    \end{itemize}
  \end{block}

@@

@@

  \begin{itemize}
  \item
    Approach: auto-generate VMs from high-level descriptions of the language
  \item
    ... using meta-programming techniques and \emph{aspects}
  \item
    high-level description: an interpreter written in a high-level language
  \item
    ... which we translate (i.e.\ compile) to VMs running on top of various
    targets, like C/Posix, CLR, JVM
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{PyPy}
  \begin{itemize}
  \item
    PyPy = Python interpreter written in RPython + translation toolchain
    for RPython
  \end{itemize}
  \pause
  \begin{block}{
    What is RPython}
    \begin{itemize}
    \item
      RPython is a subset of Python
    \item
      subset chosen in such a way that type-inference can be performed
    \item
      still a high-level language (unlike SLang or Prescheme)
    \item
      ...really a subset, can't give a small example of code that
      doesn't just look like Python :-)
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Auto-generating VMs}
  \begin{itemize}
  \item
    high-level source: early design decisions not necessary
  \item
    we need a custom \emph{translation toolchain} to compile the interpreter
    to a full VM
  \item
    many aspects of the final VM are orthogonal from the interpreter source:
    they are inserted during translation
  \item
    translation aspect $\cong$ monads, with more ad-hoc control
  \end{itemize}
  \pause
  \begin{block}{
    Examples}
    \begin{itemize}
    \item
      Garbage Collection strategy
    \item
      Threading models (e.g.\ coroutines with CPS...)
    \item
      non-trivial translation aspect: auto-generating a dynamic compiler from
      the interpreter
    \end{itemize}
  \end{block}
\end{frame}

@@

  \begin{itemize}
  \item
    dynamic languages can be implemented in a high level language
  \item
    separation of concerns from low-level details
  \item
    a potential single-source-fits-all interpreter --
    less duplication of efforts
  \item
    runs everywhere with the same semantics --
    no outdated implementations, no ties to any standard platform
  \end{itemize}
  \pause
  \begin{block}{PyPy}
     arguably the most readable Python implementation so far
  \end{block}
\end{frame}

@@

  \begin{itemize}
  \item
    ``reasonable'' performance
  \item
    can generate a dynamic compiler from the interpreter
    (work in progress, 60x faster on very simple Python code)
  \end{itemize}
  \pause
  \begin{block}{
    JIT compiler generator}
    \begin{itemize}
    \item
      almost orthogonal from the interpreter source - applicable to many
      languages, follows language evolution ``for free''
    \item
      based on Partial Evaluation
    \item
      benefits from a high-level interpreter and a tweakable translation
      toolchain
    \item
      generating a dynamic compiler is easier than generating a static one!
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Open Issues / Drawbacks / Further Work}
  \begin{itemize}
  \item
    writing the translation toolchain in the first place takes lots of effort
    (but it can be reused)
  \item
    writing a good GC is still necessary. But: maybe we can reuse existing good
    GCs (e.g.\ from the Jikes RVM)?
  \item
    conceptually simple approach but many abstraction layers
  \item
    dynamic compiler generation seems to work, but needs more efforts.
    Also: can we layer it on top of the JIT of a general purpose OO VM?
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusion / Meta-Points}
  \begin{itemize}
  \item
    high-level languages are suitable to implement dynamic languages
  \item
    doing so has many benefits
  \item
    VMs shouldn't be written by hand
  \item
    PyPy's concrete approach is not so important
  \item
    diversity is good
  \item
    let's write more meta-programming toolchains!
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Links}
  \begin{block}{
    PyPy}
    \bigskip
    \hskip 1cm \url{http://codespeak.net/pypy/}
    \bigskip
  \end{block}
\end{frame}

>>>>

\date{3rd Workshop on Dynamic Languages and Applications, July 31st 2007}

%\pgfdeclareimage[height=0.5cm]{pypy-logo}{image/py-web.png}
%\logo{\pgfuseimage{pypy-logo}}

@@

  \begin{itemize}
  \item
    implementing dynamic languages \par(with a focus on complicated ones)
  \item
    in a context of limited resources \par(academic, open source, or
    domain-specific)
  \end{itemize}
  \pause
  \begin{block}{
    Our points}
    \begin{itemize}
    \item
      Do not write virtual machines ``by hand''
    \item
      Instead, write interpreters in high-level languages
    \item
      Meta-programming is your friend
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Common Approaches to Language Implementation}
  \begin{block}{
    Using C/C++ (potentially disguised as another language)}
    \begin{itemize}
    \item
      CPython
    \item
      Ruby
    \item
      Spidermonkey (Mozilla's JavaScript VM)
    \item
      but also: Scheme48, Squeak
    \end{itemize}
  \end{block}
  \pause
  \begin{block}{
    Building on top of a general-purpose OO VM}
    \begin{itemize}
    \item
      Jython, IronPython
    \item
      JRuby, IronRuby
    \item
      various Prolog, Lisp, even Smalltalk implementations
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Implementing VMs in C}
  When writing a VM in C it is hard to reconcile:
  \begin{itemize}
  \item
    flexibility, maintainability
  \item
    simplicity
  \item
    performance
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      \alert{CPython} is a very simple bytecode VM, performance not great
    \item
      \alert{Psyco} is a just-in-time-specializer, very complex, hard to
      maintain, but good performance
    \item
      \alert{Stackless} is a fork of CPython adding microthreads. It was never
      incorporated into CPython for complexity reasons
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Fixing of Early Design Decisions}
  \begin{itemize}
  \item when starting a VM in C, many design decisions need to be made upfront
  \item examples: memory management technique, threading model
  \item such decisions are manifested throughout the VM source
  \item very hard to change later
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      CPython uses reference counting, increfs and decrefs everywhere
    \item
      CPython uses OS threads with one global lock, hard to change to
      lightweight threads or finer locking
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Compilers are a bad encoding of Semantics}
  \begin{itemize}
  \item to reach good performance levels, dynamic compilation is often needed
  \item a compiler (obviously) needs to encode language semantics
  \item this encoding is often obscure and hard to change
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      Psyco is a dynamic compiler for Python
    \item
      synchronizing with CPython's rapid development is a lot of effort
    \item
      many of CPython's new features not supported well
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Implementing Languages on Top of OO VMs}
  \begin{itemize}
  \item
    users wish to have easy interoperation with the general-purpose OO VMs used
    by the industry (JVM, CLR)
  \item
    therefore re-implementations of the language on the OO VMs are started
  \item
    more implementations!
  \item
    implementing on top of an OO VM has its own set of benefits of problems
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      \alert{Jython} is a Python-to-Java-bytecode compiler
    \item
      \alert{IronPython} is a Python-to-CLR-bytecode compiler
    \item
      both are slightly incompatible with the newest CPython version (especially
      Jython)
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Benefits of implementing on top of OO VMs}
  \begin{itemize}
  \item
    higher level of implementation
  \item
    the VM supplies a GC and mostly a JIT
  \item
    better interoperability than what the C level provides
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      both Jython and IronPython integrate well with their host OO VM
    \item
      Jython has free threading
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{The problems of OO VMs}
  \begin{itemize}
  \item
    most immediate problem: it can be hard to map
    concepts of the dynamic language to
    the host OO VM
  \item
    performance is often not improved, and can be very bad, because of the
    semantic mismatch between the dynamic language and the host VM
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      Jython about 5 times slower than CPython
    \item
      IronPython is about as fast as CPython (but some stack introspection
      features missing)
    \item
      Python has very different semantics for method calls than Java
    \end{itemize}
  \end{block}

@@

\begin{frame}
  \frametitle{Implementation Proliferation}
  \begin{itemize}
  \item
    restrictions of the original implementation lead to re-implementations,
    forks
  \item
    all implementations need to be synchronized with language evolution
  \item
    lots of duplicate effort, compatibility problems
  \end{itemize}
  \pause
  \begin{block}{
    Python Case}
    \begin{itemize}
    \item
      several serious implementations: CPython, Stackless, Psyco, Jython,
      IronPython, PyPy
    \item
      various incompatibilities
    \end{itemize}
  \end{block}
\end{frame}

@@

  \begin{itemize}
  \item
    Approach: auto-generate VMs from high-level descriptions of the language
  \item
    ... using meta-programming techniques and \emph{aspects}
  \item
    high-level description: an interpreter written in a high-level language
  \item
    ... which we translate (i.e.\ compile) to a VM running in various target
    environments, like C/Posix, CLR, JVM
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{PyPy}
  \begin{itemize}
  \item
    PyPy = Python interpreter written in RPython + translation toolchain
    for RPython
  \end{itemize}
  \pause
  \begin{block}{
    What is RPython}
    \begin{itemize}
    \item
      RPython is a subset of Python
    \item
      subset chosen in such a way that type-inference can be performed
    \item
      still a high-level language (unlike SLang or PreScheme)
    \item
      ...really a subset, can't give a small example of code that
      doesn't just look like Python :-)
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Auto-generating VMs}
  \begin{itemize}
  \item
    we need a custom \emph{translation toolchain} to compile the interpreter
    to a full VM
  \item
    many aspects of the final VM are orthogonal from the interpreter source:
    they are inserted during translation
  \item
    translation aspect $\cong$ monads, with more ad-hoc control
  \end{itemize}
  \pause
  \begin{block}{
    Examples}
    \begin{itemize}
    \item
      Garbage Collection strategy
    \item
      Threading models (e.g.\ coroutines with CPS...)
    \item
      non-trivial translation aspect: auto-generating a dynamic compiler from
      the interpreter
    \end{itemize}
  \end{block}
\end{frame}

@@

  \begin{itemize}
  \item
    dynamic languages can be implemented in a high level language
  \item
    separation of language semantics from low-level details
  \item
    a single-source-fits-all interpreter
    \begin{itemize}
      \item runs everywhere with the same semantics
      \item no outdated implementations, no ties to any standard platform
      \item less duplication of efforts
    \end{itemize}
  \end{itemize}
  \pause
  \begin{block}{PyPy}
     arguably the most readable Python implementation so far
  \end{block}
\end{frame}

@@

  \begin{itemize}
  \item
    ``reasonable'' performance
  \item
    can generate a dynamic compiler from the interpreter
    (work in progress, 60x faster on very simple Python code)
  \end{itemize}
  \pause
  \begin{block}{
    JIT compiler generator}
    \begin{itemize}
    \item
      almost orthogonal from the interpreter source – applicable to many
      languages, follows language evolution ``for free''
    \item
      based on Partial Evaluation techniques
    \item
      benefits from a high-level interpreter
    \item
      generating a dynamic compiler is easier than generating a static one!
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Drawbacks / Open Issues / Further Work}
  \begin{itemize}
  \item
    writing the translation toolchain in the first place takes lots of effort
    (but it can be reused)
  \item
    writing a good GC is still necessary. But: maybe we can reuse existing good
    GCs (e.g.\ from the Jikes RVM)?
  \item
    dynamic compiler generation seems to work, but needs more efforts.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusion / Meta-Points}
  \begin{itemize}
  \item
    VMs shouldn't be written by hand
  \item
    high-level languages are suitable to implement dynamic languages
  \item
    doing so has many benefits
  \item
    PyPy's concrete approach is not so important
  \item
    let's write more meta-programming toolchains!
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Questions?}
  \begin{block}{
    PyPy}
    \bigskip
    \hskip 1cm \url{http://codespeak.net/pypy/}
    \bigskip
  \end{block}
\end{frame}
